### 2024-04-19 docs/SOSP 2024/motivation.tex
pass

Consider a model with four trees, two 
trees of depth 1 and two of depth 2 (\circled{A} in Figure \ref{Fig:HIRExample}).
% We use this model as our running example throughout this section.
We first describe the simple schedule shown in Figure \ref{fig:schedulea} that processes one tree at a time 
for all input rows. The scheduling constructs are fairly intuitive. We 
defer a detailed explanation of the scheduling language to Section~\ref{sec:schedule}. 
The loop over the trees is split into two -- one that
iterates over the first two trees (Trees 1 and 2 with depth 1) and 
the second that iterates over the last two trees (Trees 3 and 4 with
depth 2). Straightforward traversal of trees requires a \op{while} loop
and involves branching to check if a leaf has been reached. 
One way to avoid this as done in this schedule is to unroll the 
tree walks for each tree. 
When the schedule specifies that the tree walks should be unrolled, 
\Treebeard{} pads the trees so that all leaves are at the same depth
(\circled{B} in Figure~\ref{Fig:HIRExample}).

\begin{figure*}[]
  \begin{subfigure}[t]{.495\linewidth}
    \vspace{0pt}
    \centering
    \includegraphics[width=\linewidth]{figures/MotivatingExample1.PNG}
    \caption{\label{fig:schedulea}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{.495\linewidth}
    \vspace{0pt}
    \centering
    \includegraphics[width=\linewidth]{figures/MotivatingExample2.PNG}
    \caption{\label{fig:scheduleb}}
  \end{subfigure}
  \caption{Possible schedules to compile the motivating example and the generated \Treebeard{} IR.}
  \label{Fig:Schedules}
\end{figure*}

@@

One form of parallelism that can be exploited is to process rows in 
parallel. 
%% RG:
While this may work for multi-core CPUs\footnote{As we report in Section~\ref{sec:results}, 
other ways to parallelize can benefit inference on CPUs as well.}, with massively parallel processors like GPUs,
 this strategy may not yield sufficient parallel work. To expose more parallelism, 
we can additionally parallelize across trees as done in the schedule in Figure \ref{fig:scheduleb}. 
%% RG:
% With GPUs, the performance of generated code would also vary with the 
% kernel configuration, which specifies the number of thread blocks\footnote{We 
% use NVIDIA CUDA terminology in the discussion, but as we demonstrate in 
% Section~\ref{sec:results}, our framework can generate code for AMD GPUs as well.}, 
% the size of thread blocks and so on. 
% A possible strategy to accomplish this is encoded in the following schedule.
% \begin{lstlisting}[style=c++]
%   // Split the trees into two sets
%   tile(tree, t0, t1, 2)
%   reorder(batch, t1, t0)
%   // Fiss loop so that trees with equal 
%   // depth are processed together
%   split(t0, t0_depth1, t0_depth2, 2)
%   unrollWalk(t0_depth1, 1)
%   unrollWalk(t0_depth2, 2)
%   // Configure the GPU kernel dimensions
%   gpuDimension(batch, grid.x)
%   gpuDimension(t1, block.x)
% \end{lstlisting}
This schedule generates an inference function that runs on the GPU as shown in 
the IR in Figure \ref{fig:scheduleb}. 
The inference routine processes one input row per thread block (since the \op{batch}
loop is mapped directly to \op{grid.x}).
It also splits the trees into two sets by tiling the \op{tree} loop.
Each of the two sets is processed in parallel. We unroll the tree walks 
for each tree.
% The IR generated is as follows. 
% \begin{lstlisting}[style=c++]
%   model = ensemble(...)
%   par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
%     par.for t1 = 0 to 2 step 1 <block.x> {
%       for t0_depth1 = 0 to 2 step 2 {
%         T = getTree(model, t0_depth1 + t1)
%         treePred = walkDecisionTree(T, 
%                         input[batch]) <unrollDepth = 1>
%         reduce(result[batch], treePred)
%       }
%       for t0_depth2 = 2 to 4 step 2 {
%         T = getTree(model, t0_depth2 + t1)
%         treePred = walkDecisionTree(T,
%                         input[batch]) <unrollDepth = 2>
%         reduce(result[batch], treePred) <'+', 0.0>
%       }
%     }
%   }
% \end{lstlisting}

% In the case of this schedule, the \op{reduce}
% operation needs special consideration.  
% %% RG
% %% to correctly generate code for this schedule, the compiler needs 
% To generate the correct code for this schedule, the compiler needs to determine
% the required synchronization and temporary storage for reductions 
% given the parallelization strategy specified. In the given example, it 
% has to identify that parallel iterations of the \op{t1} 
% loop accumulate into the same element of the \op{result} array.
% One possible solution is to rewrite the reduction so that each parallel 
% iteration accumulates into a different array element by introducing 
% a temporary buffer (\op{temp}) as follows.
% \begin{lstlisting}[style=c++]
%   float temp[2][BATCH_SIZE]
%   model = ensemble(...)
%   par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
%     par.for t1 = 0 to 2 step 1 <block.x> {
%       for t0_depth1 = 0 to 2 step 2 {
%         T = getTree(ensemble, t0_depth1 + t1)
%         treePred = walkDecisionTree(T, 
%                       input[batch]) <unrollDepth = 1>
%         reduce(temp[t1][batch], treePred)
%       }
%       for t0_depth2 = 2 to 4 step 2 {
%         T = getTree(ensemble, t0_depth2 + t1)
%         treePred = walkDecisionTree(T,
%                       input[batch]) <unrollDepth = 2>
%         reduce(temp[t1][batch], treePred) <'+', 0.0>
%       }
%     }
%     result[batch] = reduce_dimension(temp[:][batch], 0)
%   }
% \end{lstlisting}
% Here, partial results are accumulated into \op{temp} and then
% reduced across the \op{t1} dimension to get the final result.
Note that while unrolling helps avoid branching, it likely increases 
the total amount of computation. Another option is to not unroll 
and let the GPU manage the branching. The schedules with and 
without unrolling place different constraints on the target processor,
and the best choice depends on the characteristics of the model
and micro-architectural features like register file size, handling 
of branch divergence etc.~\cite{FungMicro, MilindDivergence}.

% As is evident from these examples, it is possible to optimize 
% the inference routine in different ways. Also, the structure of the loop 
% nest in the inference routine can get quite complex even 
% for simple schedules. Writing these routines by hand 
% is error-prone and time-consuming.  
% %% RG 
% %% We beleive 
% Hence designing a 
% scheduling language to encapsulate these strategies and a principled code 
% generator to automatically generate code guided by the schedule is the 
% best approach. We promote such an approach in this paper.

>>>>

Consider a model with four trees, two 
trees of depth 1 and two of depth 2 (\circled{A} in Figure \ref{Fig:HIRExample}).
% We use this model as our running example throughout this section.
We first describe the simple schedule shown in Figure \ref{fig:schedulea} that processes one tree at a time 
for all input rows. The scheduling constructs are fairly intuitive. We 
defer a detailed explanation of the scheduling language to Section~\ref{sec:schedule}. 
In this schedule, the loop over the trees is split into two -- one that
iterates over the first two trees (Trees 1 and 2 with depth 1) and 
the second that iterates over the last two trees (Trees 3 and 4 with
depth 2). Straightforward traversal of trees requires a \op{while} loop
and involves branching to check if a leaf has been reached. 
One way to avoid this, as done in this schedule, is to unroll the 
tree walks for each tree. 
When the schedule specifies that the tree walks should be unrolled, 
\Treebeard{} pads the trees so that all leaves are at the same depth
(\circled{B} in Figure~\ref{Fig:HIRExample}).

\begin{figure*}[]
  \begin{subfigure}[t]{.495\linewidth}
    \vspace{0pt}
    \centering
    \includegraphics[width=\linewidth]{figures/MotivatingExample1.PNG}
    \caption{\label{fig:schedulea}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{.495\linewidth}
    \vspace{0pt}
    \centering
    \includegraphics[width=\linewidth]{figures/MotivatingExample2.PNG}
    \caption{\label{fig:scheduleb}}
  \end{subfigure}
  \caption{Some possible schedules to compile the motivating example and the generated \Treebeard{} IR.}
  \label{Fig:Schedules}
\end{figure*}

@@

One form of parallelism that can be exploited is to process rows in 
parallel. 
%% RG:
While this may work for multi-core CPUs\footnote{As we report in Section~\ref{sec:results}, 
other ways to parallelize can benefit inference on CPUs as well.}, with massively parallel processors like GPUs,
 this strategy may not yield sufficient parallel work. To expose more parallelism, 
we can additionally parallelize across trees as done in the schedule in Figure \ref{fig:scheduleb}. 
%% RG:
% With GPUs, the performance of generated code would also vary with the 
% kernel configuration, which specifies the number of thread blocks\footnote{We 
% use NVIDIA CUDA terminology in the discussion, but as we demonstrate in 
% Section~\ref{sec:results}, our framework can generate code for AMD GPUs as well.}, 
% the size of thread blocks and so on. 
% A possible strategy to accomplish this is encoded in the following schedule.
% \begin{lstlisting}[style=c++]
%   // Split the trees into two sets
%   tile(tree, t0, t1, 2)
%   reorder(batch, t1, t0)
%   // Fiss loop so that trees with equal 
%   // depth are processed together
%   split(t0, t0_depth1, t0_depth2, 2)
%   unrollWalk(t0_depth1, 1)
%   unrollWalk(t0_depth2, 2)
%   // Configure the GPU kernel dimensions
%   gpuDimension(batch, grid.x)
%   gpuDimension(t1, block.x)
% \end{lstlisting}
The corresponding GPU inference routine is shown in the same figure. 
This schedule generates a routine that processes one input row per thread block (\op{batch}
loop is mapped to \op{grid.x}).
The schedule also tiles the \op{tree} loop into a nest of two loops with indices $t0$ and $t1$. 
It then additionally parallelizes across the $t1$ loop.
Finally, it splits $t0$ and unrolls the walks for each tree depth.
%It also splits the trees into two sets by tiling the \op{tree} loop.
%Each of the two sets is processed in parallel. We unroll the tree walks 
%for each tree.
% The IR generated is as follows. 
% \begin{lstlisting}[style=c++]
%   model = ensemble(...)
%   par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
%     par.for t1 = 0 to 2 step 1 <block.x> {
%       for t0_depth1 = 0 to 2 step 2 {
%         T = getTree(model, t0_depth1 + t1)
%         treePred = walkDecisionTree(T, 
%                         input[batch]) <unrollDepth = 1>
%         reduce(result[batch], treePred)
%       }
%       for t0_depth2 = 2 to 4 step 2 {
%         T = getTree(model, t0_depth2 + t1)
%         treePred = walkDecisionTree(T,
%                         input[batch]) <unrollDepth = 2>
%         reduce(result[batch], treePred) <'+', 0.0>
%       }
%     }
%   }
% \end{lstlisting}

% In the case of this schedule, the \op{reduce}
% operation needs special consideration.  
% %% RG
% %% to correctly generate code for this schedule, the compiler needs 
% To generate the correct code for this schedule, the compiler needs to determine
% the required synchronization and temporary storage for reductions 
% given the parallelization strategy specified. In the given example, it 
% has to identify that parallel iterations of the \op{t1} 
% loop accumulate into the same element of the \op{result} array.
% One possible solution is to rewrite the reduction so that each parallel 
% iteration accumulates into a different array element by introducing 
% a temporary buffer (\op{temp}) as follows.
% \begin{lstlisting}[style=c++]
%   float temp[2][BATCH_SIZE]
%   model = ensemble(...)
%   par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
%     par.for t1 = 0 to 2 step 1 <block.x> {
%       for t0_depth1 = 0 to 2 step 2 {
%         T = getTree(ensemble, t0_depth1 + t1)
%         treePred = walkDecisionTree(T, 
%                       input[batch]) <unrollDepth = 1>
%         reduce(temp[t1][batch], treePred)
%       }
%       for t0_depth2 = 2 to 4 step 2 {
%         T = getTree(ensemble, t0_depth2 + t1)
%         treePred = walkDecisionTree(T,
%                       input[batch]) <unrollDepth = 2>
%         reduce(temp[t1][batch], treePred) <'+', 0.0>
%       }
%     }
%     result[batch] = reduce_dimension(temp[:][batch], 0)
%   }
% \end{lstlisting}
% Here, partial results are accumulated into \op{temp} and then
% reduced across the \op{t1} dimension to get the final result.
Note that while unrolling helps avoid branching, it increases 
the total amount of computation. Another option is to not unroll 
and let the GPU manage the branching. The schedules with and 
without unrolling place different constraints on the target processor,
and the best choice depends on the characteristics of the model
and micro-architectural features like register file size and handling 
of branch divergence~\cite{FungMicro, MilindDivergence}.
These schedules just use a combination of 4-8 constructs and already generate schedules 
that are different from what library based systems like XGBoost, Tahoe and RAPIDs FIL use. 
As one can imagine several other schedules with different trade-offs can be generating using the constructs.
% As is evident from these examples, it is possible to optimize 
% the inference routine in different ways. Also, the structure of the loop 
% nest in the inference routine can get quite complex even 
% for simple schedules. Writing these routines by hand 
% is error-prone and time-consuming.  
% %% RG 
% %% We beleive 
% Hence designing a 
% scheduling language to encapsulate these strategies and a principled code 
% generator to automatically generate code guided by the schedule is the 
% best approach. We promote such an approach in this paper.
